---
title: "SPEIChanges: Assessing Changes in Drought Frequency in Oxford, UK"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
This vignette demonstrates how to use the `SPEIChanges` package to assess changes in drought frequency in Oxford, UK. We will analyze daily air temperature and rainfall data from the Radcliffe Observatory, covering the period from January 1827 to January 2020. The vignette serves as a workflow for applying the `SPEIChanges` method and interpreting its results.

## Getting Started

Load the library in your R session.

```{r setup}
library(SPEIChanges)
```

## Data
The SPEIChanges method is based on the SPEI (Standardized Precipitation Evapotranspiration Index), which requires precipitation and evapotranspiration data. In this vignette, we will use the `Oxford` dataset, which contains daily air temperature and rainfall data from the Radcliffe Observatory in Oxford, UK.
It includes columns for years, months, days, maximum and minimum daily air temperatures, and rainfall.

```{r data}
data("Oxford")
head(Oxford)
```

## Transforming daily data into monthly P - PE values
```{r transform}
library(lubridate)
library(dplyr)
library(SPEI)
lat <- 51.75 # Latitude of Oxford, UK
monthly_Oxford <- Oxford %>%
  group_by(YYYY, MM) %>%
  summarise(
    Tmax     = mean(Tmax, na.rm = TRUE),
    Tmin     = mean(Tmin, na.rm = TRUE),
    Rainfall = sum(Rainfall, na.rm = TRUE),
    .groups = "drop"
  )

Tavg <- (monthly_Oxford$Tmin + monthly_Oxford$Tmax)/2
ts.Tavg <- ts(Tavg,frequency = 12,start = c(1827,1))
PE <- thornthwaite(Tave=ts.Tavg, lat=lat, na.rm = FALSE, verbose=TRUE)
PE <- as.numeric(PE)
PPE <- monthly_Oxford$Rainfall - PE
```

## Aggregating P - PE values into three time scales (1-, 3- and 6-months)

```{r aggregate}
# 1-month time scale
PPE_atTS1 <- PPEaggreg_month(monthly.PPE=PPE, 
                start.year=1827, 
                start.month = 1, 
                TS = 1)
# 3-month time scale
PPE_atTS3 <- PPEaggreg_month(monthly.PPE=PPE,
                start.year=1827,
                start.month = 1,
                TS = 3)
# 6-month time scale
PPE_atTS6 <- PPEaggreg_month(monthly.PPE=PPE,
                start.year=1827,
                start.month = 1,
                TS = 6)
```

## Assessing changes in drought frequency using the `SPEIChanges` method
```{r changes}
# 1-month time scale
Changes_SPEI_1month <- SPEIChanges_month(PPE.at.TS=PPE_atTS1, 
                  nonstat.models = 3, 
                  criterion = "BIC")
# 3-month time scale
Changes_SPEI_3month <- SPEIChanges_month(PPE.at.TS=PPE_atTS3,
                  nonstat.models = 3, 
                  criterion = "BIC")
# 6-month time scale
Changes_SPEI_6month <- SPEIChanges_month(PPE.at.TS=PPE_atTS6, 
                  nonstat.models = 3, 
                  criterion = "BIC")
```

## Interpreting the results

The `SPEIChanges_month` function returns a list with: $Changes.Freq.Drought, $data.month, and $GEV.parameters. \


- **$Changes.Freq.Drought:** This object contains the non-stationary GEV-based model that best fits the P - PE series, the expected P - PE amounts under both stationary and non-stationary approaches for normal conditions, and the frequency changes for moderate-to-extreme, severe-to-extreme, and extreme drought events. It has 12 rows (one for each month) and 6 columns. This is the package's most important output, as it identifies months where drought occurrence changes have been detected and specifies the non-stationary model describing these changes. \

- **$data.month:** This object contains the P - PE amounts, SPEI values, cumulative probabilities under both stationary and non-stationary approaches, and the frequency changes in below-zero SPEI values caused by shifts in P - PE patterns. It has the same number of rows as the input data. \

- **$GEV.parameters:** This object contains the time-varying parameters of the best-fitting GEV model (location, scale, and shape). If the input data has 30 years of records, this object will have 360 rows (30 years x 12 months) and 5 columns (month, year, location, scale, shape). This output is useful for understanding how the parameters of the GEV distribution evolve over time for each month. \

### Detailed look at the $Changes.Freq.Drought object: 1-month time scale

The `$Changes.Freq.Drought` object for the 1-month time scale shows the changes in drought frequency by month:

```{r display_changes_freq, results='asis'}
knitr::kable(Changes_SPEI_1month$Changes.Freq.Drought,
             caption = "Changes in drought frequency by month (1-month time scale)",
             digits = 4)
```

- When applied at the 1-month time scale, the **$Changes.Freq.Drought** object indicates increasing drought frequency during the summer (June and July) and autumn (September and October) seasons. These changes are expressed as percentages in the columns *ChangeMod*, *ChangeSev*, and *ChangeExt*, which represent shifts in the frequency of moderate-to-extreme, severe-to-extreme, and extreme drought events, respectively. During these months, decreases were also observed in the expected climatic P–PE values (i.e., the P–PE values corresponding to SPEI = 0). Specifically, the columns *StatNormalPPE* and *NonStatNormalPPE* provide these expected values under stationary and non-stationary assumptions, respectively. In contrast, December exhibits the opposite behavior, with changes indicating a decrease in drought frequency.

### Detailed look at the $data.month object: 1-month time scale

The `$data.month` object for the 1-month time scale contains the following data:

```{r display_data_month, results='asis'}
knitr::kable(Changes_SPEI_1month$data.month, 
             caption = "Monthly data with stationary and non-stationary probabilities (1-month time scale)",
             digits = 4)
```


- As can be noted in the table above, the `$data.month` object includes the P - PE values, SPEI values, cumulative probabilities under both stationary and non-stationary approaches, and the frequency changes in below-zero SPEI values caused by shifts in P - PE patterns. This data allows us to analyze how the probability of drought conditions (SPEI < 0) has evolved over time for each month, and how it differs between stationary and non-stationary assumptions.

### $GEV.parameters

The `$GEV.parameters` object for the 1-month time scale contains the time-varying GEV parameters:

```{r display_gev_params, results='asis'}
knitr::kable(Changes_SPEI_1month$GEV.parameters,
             caption = "Time-varying GEV parameters by month and year (1-month time scale)",
             digits = 4)
```

- As shown in the table above, the `$GEV.parameters` object includes the month and the location, scale, and shape parameters of the best-fitting GEV model for each month and year. This information is crucial for understanding how the distribution of P - PE values changes over time and how it affects drought frequency.
