---
title: "SPEIChanges: Assessing Changes in Drought Frequency in Oxford, UK"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
This vignette demonstrates how to use the `SPEIChanges` package to assess changes in drought frequency in Oxford, UK. We will analyze daily air temperature and rainfall data from the Radcliffe Observatory, covering the period from January 1827 to January 2020. The vignette serves as a workflow for applying the `SPEIChanges` method and interpreting its results.

## Getting Started

Load the library in your R session.

```{r setup}
library(SPEIChanges)
```

## Data
The SPEIChanges method is based on the SPEI (Standardized Precipitation Evapotranspiration Index), which requires precipitation and evapotranspiration data. In this vignette, we will use the `Oxford` dataset, which contains daily air temperature and rainfall data from the Radcliffe Observatory in Oxford, UK.
It includes columns for years, months, days, maximum and minimum daily air temperatures, and rainfall.

```{r data}
data("Oxford")
head(Oxford)
```

## Transforming daily data into monthly P - PE values
```{r transform}
library(lubridate)
library(dplyr)
library(SPEI)
lat <- 51.75 # Latitude of Oxford, UK
monthly_Oxford <- Oxford %>%
  group_by(YYYY, MM) %>%
  summarise(
    Tmax     = mean(Tmax, na.rm = TRUE),
    Tmin     = mean(Tmin, na.rm = TRUE),
    Rainfall = sum(Rainfall, na.rm = TRUE),
    .groups = "drop"
  )

Tavg <- (monthly_Oxford$Tmin + monthly_Oxford$Tmax)/2
ts.Tavg <- ts(Tavg,frequency = 12,start = c(1827,1))
PE <- thornthwaite(Tave=ts.Tavg, lat=lat, na.rm = FALSE, verbose=TRUE)
PE <- as.numeric(PE)
PPE <- monthly_Oxford$Rainfall - PE
```

## Aggregating P - PE values into three time scales (1-, 3- and 6-months)

```{r aggregate}
# 1-month time scale
PPE_atTS1 <- PPEaggreg_month(monthly.PPE=PPE, 
                start.year=1827, 
                start.month = 1, 
                TS = 1)
# 3-month time scale
PPE_atTS3 <- PPEaggreg_month(monthly.PPE=PPE,
                start.year=1827,
                start.month = 1,
                TS = 3)
# 6-month time scale
PPE_atTS6 <- PPEaggreg_month(monthly.PPE=PPE,
                start.year=1827,
                start.month = 1,
                TS = 6)
```

## Assessing changes in drought frequency using the `SPEIChanges` method
```{r changes}
# 1-month time scale
Changes_SPEI_1month <- SPEIChanges_month(PPE.at.TS=PPE_atTS1, 
                  nonstat.models = 3, 
                  criterion = "BIC")
# 3-month time scale
Changes_SPEI_3month <- SPEIChanges_month(PPE.at.TS=PPE_atTS3,
                  nonstat.models = 3, 
                  criterion = "BIC")
# 6-month time scale
Changes_SPEI_6month <- SPEIChanges_month(PPE.at.TS=PPE_atTS6, 
                  nonstat.models = 3, 
                  criterion = "BIC")
```
## Interpreting the results

                                         
